---
title: "The Relationship between Financial and Job Satisfaction and Levels of Happiness Among Different Social Classes "
author: "Carrie Magee, Abhimanyu Barun, Ambar Pathak, and Kismat Khatri"
output: html_document
---

```{r setup, include=FALSE}
#uploading packages
library(ezids)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
gss1 <- data.frame(read.csv("/Users/carriemagee/Desktop/GROUP2_DATS6101_MIDTERM_PROJECT/GSS.csv"))
#cleaning data
gss1_h = subset(gss1, happy == "Very happy" | happy == "Pretty happy" | happy == "Not too happy")
gss1_hsj = subset(gss1_h, satjob == "Very satisfied" | satjob == "Moderately satisfied" | satjob == "A little dissatisfied" | satjob == "Very dissatisfied")
gss1_hsjc = subset(gss1_hsj, class_ == "Lower class" | class_ == "Working class" | class_ == "Middle class" | class_ == "Upper class")
gss1_hsjcsf = subset(gss1_hsjc, satfin == "Pretty well satisfied" | satfin == "More or less satisfied" | satfin == "Not satisfied at all")
```


## Introduction

  There is still contention over whether money and success are the key to real happiness. Americans are often forced to sacrifice their happiness due to a culture that puts immense focus on establishing a successful career and working towards financial freedom. Furthermore, studies have shown the potential effects social class has on individual’s thoughts, feelings, and behaviors (Manstead, 2018). In relation, individuals of different social classes may have varying life experiences that can influence their financial and job satisfaction as well as their general happiness. 
  
  With the use of data from the General Social Survey, our report focuses on the general happiness of a individual in relation to their work satisfaction and financial satisfaction from 1980 to 2018. In order to measure job satisfaction, we used the statement *On the whole, how satisfied are you with the work you do..?* with responses ranging from *very satisfied, moderately satisfied, a little dissatisfied, or very dissatisfied*. Similarly, financial satisfaction was measured using the question *So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?* Our third variable "general happiness" was measured by asking *Taken all together, how would you say things are these days--would you say that you are very happy, pretty happy, or not too happy?*
  
  More specifically, we hoped to uncover any potential relationships between an individual's social class and their job satisfaction,financial satisfaction, and general happiness. Social class was measured by asking *If you were asked to use one of four names for your social class, which would you say you belong in?* with four responses including *the lower class, the working class, the middle class, or the upper class*. 
  

## General Social Survey dataset

### Descriptive Statistics
```{r}
pie_happy <- gss1_hsjcsf %>% count(happy) %>%
  arrange(desc(happy)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

ggplot(data= pie_happy)+
  geom_col(mapping = aes(x="", y=percentages, fill=happy)) +
  coord_polar(theta = "y") +
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))
```
The Pie chart shows the more than half of the people are pretty happy from 1972 to 2021.




```{r}
pie_class <- gss1_hsjcsf %>% count(class_) %>%
  arrange(desc(class_)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

ggplot(data= pie_class)+
  geom_col(mapping = aes(x="", y=percentages, fill=class_)) +
  coord_polar(theta = "y") +
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))
```

 A pie chart shows the classes of people of America. It show that nearly half of the people fall under working class and about 43% people are middle class in America.
 
```{r}
pie_satfin <- gss1_hsjcsf %>% count(satfin) %>%
  arrange(desc(satfin)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

ggplot(data= pie_satfin)+
  geom_col(mapping = aes(x="", y=percentages, fill=satfin)) +
  coord_polar(theta = "y") +
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))
```


It shows that nearly half of the people are more or less satisfied with their financial status whereas about same percentage of people are pretty well satisfied and Not satisfied at all with their Financial.



```{r}
pie_satjob <- gss1_hsjcsf %>% count(satjob) %>%
  arrange(desc(satjob)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

ggplot(data= pie_satjob)+
  geom_col(mapping = aes(x="", y=percentages, fill=satjob)) +
  coord_polar(theta = "y") +
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))

```
The pie chart shows about 48% of american people's are very satisfied with their job whereas a little people are dissatisfied with their job.

INSERT COMMENTARY HERE

### Chi Squared Analysis
```{r}
sjPlot::tab_xtab(var.row = gss1_hsjcsf$happy, var.col =gss1_hsjcsf$satfin, title = "Level of Happiness by Financial Satisfaction Crosstabulation",var.labels = c("Happiness Level","Financial Satisfaction"), show.row.prc = TRUE,emph.total = TRUE)
```

The cross tabulation above shows the frequency of happiness levels by financial satisfaction. We wanted to further examine the relationship between financial satisfaction and happiness in order to see how feelings of financial satisfaction can potentially influence how happy an individual feels in life.  According to the chi-squared test of independence, there is a significant relationship between happiness and financial satisfaction (p<0.05). This means that an individual's level of happiness depends on their feelings of satisfaction with their current financial situation. More specifically, the table shows that a majority of individuals (54.5%) that report being not at all satisfied with their financial situation were not too happy in life. In addition, almost 49% of people that reported being more or less satisfied with their financial situation were pretty happy with their life while about 42% of individuals that were pretty well satisfied with their financial situation reported being very happy in life. The results clearly imply that the more financially satisfied an individual is, the happier they are in life. 

```{r fig.height = 8, fig.width = 13}
sjPlot::tab_xtab(var.row = gss1_hsjcsf$happy, var.col = gss1_hsjcsf$satjob,title = "Level of Happiness by Job Satisfaction Crosstabulation",var.labels = c("Happiness Level","Job Satisfaction"), show.row.prc = TRUE,emph.total = TRUE)
```

The cross tabulation above shows the frequency of happiness levels by job satisfaction. We wanted to further examine the relationship between job satisfaction and happiness in order to see how job satisfaction can potentially influence how happy an individual feels in life. According to the chi-squared test of independence, there is a significant relationship between job satisfaction and happiness (p<0.05). This means that an individual's level of happiness depends on their satisfaction with their current job. The table shows that individuals that a majority of individuals that reported being very dissatisfied (12.2%) or a little dissatisfied (19.8%) with their jobs were not too happy in their lives. About 44% of people that were moderately satisfied with their jobs reported being pretty happy in life while 65.9% of people who were very satisfied with their jobs reported being very happy in life. Considering that individuals who reported being more dissatisfied with job also reported being unhappy and those who reported being more satisfied were more happy shows  that job satisfaction may play a crucial role in how happy individuals are in life. 

### Graph 1
```{r fig.height = 8, fig.width = 13}
# making plot of relationship between class and job satisfaction
ggplot(gss1_hsjcsf, aes(x=satjob, fill=happy)) +
  geom_bar(position = "dodge")+
  facet_wrap(~class_)  +
  labs(title="Relationship Between Job Satisfaction and Happiness Levels by Social Class",x="Job Satisfaction", y = "Frequency",fill="Level of Happiness")+
  theme_bw()+theme(plot.title = element_text(face="bold",hjust = 0.5))+
  scale_fill_brewer(palette = "Set2")+scale_x_discrete(limits = c("Very dissatisfied","A little dissatisfied", "Moderately satisfied","Very satisfied"))
```

The figure above looks at the relationship between individuals' job satisfaction and general happiness by social class. For context, there is a larger number of working and middle class individuals than upper and lower class individuals. The results show that many lower class individuals were very satisfied with their job and that those same individuals were pretty happy overall. A majority of working class individuals were also very satisfied with their job and were also pretty happy. Interestingly, many middle class individuals were very satisfied with their jobs but reported being equally pretty happy and very happy in their lives. The upper class individuals were the only group that had more very happy individuals with very high job satisfaction. In other words, most individuals within the four social class groups were very satisfied with their jobs but only upper class individuals had more very happy individuals compared to pretty happy individuals. These results provide some insight into how social class may have a greater influence on individual's feelings of financial satisfaction and thus also influencing their feelings of happiness in life.  

### Graph 2
```{r fig.height = 7, fig.width = 12}
# making plot of relationship between class and financial satisfaction
ggplot(gss1_hsjcsf, aes(x=satfin, fill=happy)) +
  geom_bar(position = "dodge")+
  facet_wrap(~class_)  +
  labs(title="Relationship Between Financial Satisfaction and Happiness Levels by Social Class",x="Financial Satisfaction", y = "Frequency",fill="Level of Happiness")+
  theme_bw()+theme(plot.title = element_text(face="bold",hjust = 0.5))+
  scale_fill_brewer(palette = "Set2")+scale_x_discrete(limits = c("  Not satisfied at all","More or less satisfied","Pretty well satisfied"),guide = guide_axis(n.dodge=2))
```

This figure illustrates the relationship between individuals' financial satisfaction and general happiness by social class. There is also a disproportionate number of individuals in the middle and working class groups in comparison to the lower and upper class individuals. For lower, working, and middle class groups, a majority of individuals were more or less satisfied with their financial situation and reported feeling pretty happy in their life. It is important to note that there is a similar trend in the middle class group for both job satisfaction and financial satisfaction in relation to happiness. Middle class individuals that are highly satisfied with their job or financial situation were equally high in both the pretty happy and very happy responses.For upper class individuals, a large number were pretty well satisfied with their financial situation and reported being very happy in life. As we can see, the results show that the higher the social class the more you become satisfied with your financial situation and the more you report being very happy while the more you go down in class, the more your satisfaction and happiness become average (as in more or less satisfied and pretty happy).

### Pie Chart
```{r}
library(grid)
library(gridExtra)
par(mfrow=c(1,4))
pie_happy <- gss1_hsjcsf %>% count(happy) %>%
  arrange(desc(happy)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

pie_class <- gss1_hsjcsf %>% count(class_) %>%
  arrange(desc(class_)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

pie_satfin <- gss1_hsjcsf %>% count(satfin) %>%
  arrange(desc(satfin)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

pie_satjob <- gss1_hsjcsf %>% count(satjob) %>%
  arrange(desc(satjob)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

plot1<-ggplot(pie_happy, aes(x="", y=percentages, fill=happy)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)

plot2<-ggplot(pie_class, aes(x="", y=percentages, fill=class_)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)

plot3<-ggplot(pie_satfin, aes(x="", y=percentages, fill=satfin)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)

plot4<-ggplot(pie_satjob, aes(x="", y=percentages, fill=satjob)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)

grid.arrange(plot1, plot2, plot3, plot4, ncol=3)
```
```

### Time-Series Analysis
```{r}

```


INSERT COMMENTARY
