---
title: "The Relationship between Financial and Job Satisfaction and Levels of Happiness Among Different Social Classes "
author: "Carrie Magee, Abhimanyu Barun, Ambar Pathak, and Kismat Khatri"
output: html_document
---

```{r setup, include=FALSE}
#uploading packages
library(ezids)
library(dplyr)
library(ggplot2)
library(plotly)
library(ggeasy)
library(ggpubr)
library(gganimate)
library(gifski)
library(transformr)
library(scales)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
gss1 <- data.frame(read.csv("/Users/carriemagee/Desktop/GROUP2_DATS6101_MIDTERM_PROJECT/GSS.csv"))
#cleaning data
gss1_h = subset(gss1, happy == "Very happy" | happy == "Pretty happy" | happy == "Not too happy")
gss1_hsj = subset(gss1_h, satjob == "Very satisfied" | satjob == "Moderately satisfied" | satjob == "A little dissatisfied" | satjob == "Very dissatisfied")
gss1_hsjc = subset(gss1_hsj, class_ == "Lower class" | class_ == "Working class" | class_ == "Middle class" | class_ == "Upper class")
gss1_hsjcsf = subset(gss1_hsjc, satfin == "Pretty well satisfied" | satfin == "More or less satisfied" | satfin == "Not satisfied at all")
```


## Introduction

  There is still contention over whether money and success are the key to real happiness. Americans are often forced to sacrifice their happiness due to a culture that puts immense focus on establishing a successful career and working towards financial freedom. Furthermore, studies have shown the potential effects social class has on individual’s thoughts, feelings, and behaviors (Manstead, 2018). In relation, individuals of different social classes may have varying life experiences that can influence their financial and job satisfaction as well as their general happiness. 
  
  With the use of data from the General Social Survey, our report focuses on the general happiness of a individual in relation to their work satisfaction and financial satisfaction from 1980 to 2018. In order to measure job satisfaction, we used the statement *On the whole, how satisfied are you with the work you do..?* with responses ranging from *very satisfied, moderately satisfied, a little dissatisfied, or very dissatisfied*. Similarly, financial satisfaction was measured using the question *So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?* Our third variable "general happiness" was measured by asking *Taken all together, how would you say things are these days--would you say that you are very happy, pretty happy, or not too happy?*
  
  More specifically, we hoped to uncover any potential relationships between an individual's social class and their job satisfaction,financial satisfaction, and general happiness. Social class was measured by asking *If you were asked to use one of four names for your social class, which would you say you belong in?* with four responses including *the lower class, the working class, the middle class, or the upper class*. 
  

## General Social Survey dataset

### Descriptive Statistics
```{r}

```

INSERT COMMENTARY HERE

### Graph 1
```{r fig.height = 8, fig.width = 13}
# making plot of relationship between class and job satisfaction
ggplot(gss1_hsjcsf, aes(x=satjob, fill=happy)) +
  geom_bar(position = "dodge")+
  facet_wrap(~class_)  +
  labs(title="Relationship Between Job Satisfaction and Happiness Levels by Social Class",x="Job Satisfaction", y = "Frequency",fill="Level of Happiness")+
  theme_bw()+theme(plot.title = element_text(face="bold",hjust = 0.5))+
  scale_fill_brewer(palette = "Set2")+scale_x_discrete(limits = c("Very dissatisfied","A little dissatisfied", "Moderately satisfied","Very satisfied"))
```

The figure above looks at the relationship between individuals' job satisfaction and general happiness by social class. For context, there is a larger number of working and middle class individuals than upper and lower class individuals. The results show that many lower class individuals were very satisfied with their job and that those same individuals were pretty happy overall. A majority of working class individuals were also very satisfied with their job and were also pretty happy. Interestingly, many middle class individuals were very satisfied with their jobs but were also equally pretty happy and very happy. The upper class individuals were the only group that had more very happy individuals with high job satisfaction. In other words, most individuals within the four social class groups were very satisfied with their jobs but only upper class individuals had more very happy individuals compared to pretty happy individuals.  

### Graph 2
```{r fig.height = 7, fig.width = 12}
# making plot of relationship between class and financial satisfaction
ggplot(gss1_hsjcsf, aes(x=satfin, fill=happy)) +
  geom_bar(position = "dodge")+
  facet_wrap(~class_)  +
  labs(title="Relationship Between Financial Satisfaction and Happiness Levels by Social Class",x="Financial Satisfaction", y = "Frequency",fill="Level of Happiness")+
  theme_bw()+theme(plot.title = element_text(face="bold",hjust = 0.5))+
  scale_fill_brewer(palette = "Set2")+scale_x_discrete(limits = c("  Not satisfied at all","More or less satisfied","Pretty well satisfied"),guide = guide_axis(n.dodge=2))
```

This figure illustrates the relationship between individuals' financial satisfaction and general happiness by social class. There is also a disproportionate number of individuals in the middle and working class groups in comparison to the lower and upper class individuals. For lower, working, and middle class groups, a majority of individuals were more or less satisfied with their financial situation and reported feeling pretty happy in their life. It is important to note that there is a similar trend in the middle class group for both job satisfaction and financial satisfaction in relation to happiness. Individuals that are highly satisfied with their job or financial situation were equally high in both pretty happiness and very happiness.For upper class individuals, a large number were pretty well satisfied with their financial situation and reported being very happy in life. 


### Chi Squared Analysis
```{r}
sjPlot::tab_xtab(var.row = gss1_hsjcsf$satfin, var.col =gss1_hsjcsf$class_, title = "Financial Satisfaction by Social Class Crosstabulation",var.labels = c("Financial Satisfaction","Social Class"), show.row.prc = TRUE,emph.total = TRUE)
```

The cross tabulation above shows the frequency of financial satisfaction levels by social class. According to the chi-squared test of independence, there is a significant relationship between financial satisfaction and social class (p<0.05). This means that an individual's level of satisfaction with their financial situation depends on their social class. More specifically, the table shows that a majority of lower class individuals (12.5%) were not at all satisfied with their financial situations. Similarly, almost 70% of working class individuals were not satisfied at all with their financial situation. In contrast, a majority of middle class (59.3%) and upper class (6.8%) individuals were pretty well satisfied with their financial situation. Overall, the results show a large difference in satisfaction with financial situations across social classes.

```{r fig.height = 8, fig.width = 13}
sjPlot::tab_xtab(var.row = gss1_hsjcsf$satjob, var.col = gss1_hsjcsf$class_,title = "Job Satisfaction by Social Class Crosstabulation",var.labels = c("Job Satisfaction ","Social Class"), show.row.prc = TRUE,emph.total = TRUE)
```

The cross tabulation above shows the frequency of job satisfaction levels by social class.According to the chi-squared test of independence, there is a significant relationship between financial satisfaction and social class (p<0.05). This means that an individual's level of satisfaction with their job depends on their social class. The results show that most lower class individuals (14.9%) are very dissatisfied with their jobs while 53.7% of working class individuals are a little dissatisfied with their job. In contrast, a majority of middle class (47.4%) and upper class (4%) individuals are very satisfied with their jobs. Nonetheless, the results reveal a divide between lower and working class individuals and middle and upper class individuals in relation to their overall job satisfaction. 


### Time-Series Analysis

  We created one subset of dataframe by adding a column of  percentage value 
  of every possible combination of happiness, year and job satisfaction and similarly
  another subset for that showed percentage value of every possible combination of 
  happiness, year and financial satisfaction.

```{r}

# Calculations prior to plotting happiness levels compared to the job satisfaction over the course of 50 Years.

gss_satjob <- gss1_hsjcsf %>% 
  group_by(year, happy, satjob) %>% 
  summarise(total_count=n(),
            .groups = 'drop') %>% 
  as.data.frame()

gss_satjob_perc <- transform(gss_satjob,
                             perc = ave(total_count,
                                        year,
                                        FUN = prop.table))

gss_satjob_perc$perc = gss_satjob_perc$perc * 100


# Plot to analyze happiness levels compared to the job satisfaction over the course of 50 Years.

ggplotly(gss_satjob_perc %>%
           ggplot( aes(x=year, y=perc)) +
           geom_line(aes(colour=happy)) +
           geom_point()+
           facet_wrap(. ~ satjob, nrow = 4)+
           ylab("Percentage Happiness ")+
           scale_color_manual(values=c("#19272e", "#c7a04e", "#39db7f")))


gif1 <- gss_satjob_perc %>%
           ggplot( aes(x=year, y=perc)) +
           geom_line(aes(colour=happy)) +
           geom_point()+
           ylab("Percentage Happiness ")+
           facet_wrap(. ~ satjob, nrow = 4)+
           transition_reveal(year)+
           scale_color_manual(values=c("#19272e", "#c7a04e", "#39db7f"))


animate(gif1, renderer = gifski_renderer(), fps = 60, height = 690, width =1000, duration = 3)
anim_save("gif1.gif")

```
This Plot illustrates that the percentage happiness changes with time and the financial and socio economic situation at a certain point of time in history till date. 

```{r}

# Calculations prior to plotting happiness levels compared to the financial satisfaction over the course of 50 Years.

gss_satfin <- gss1_hsjcsf %>% 
  group_by(year, happy, satfin) %>% 
  summarise(total_count=n(),
            .groups = 'drop') %>% 
  as.data.frame()

gss_satfin_perc <- transform(gss_satfin,
                             perc = ave(total_count,
                                        year,
                                        FUN = prop.table))

gss_satfin_perc$perc = gss_satfin_perc$perc * 100

# Plot to analyze happiness levels compared to the financial satisfaction over the course of 50 Years.

ggplotly(gss_satfin_perc %>%
           ggplot( aes(x=year, y=perc)) +
           geom_line(aes(colour=happy)) +
           geom_point()+
           facet_wrap(. ~ satfin, nrow = 3)+
           ylab("Percentage Happiness ")+
           scale_color_manual(values=c("#19272e", "#c7a04e", "#39db7f")))  

gif2<-gss_satfin_perc %>%
  ggplot( aes(x=year, y=perc)) +
  geom_line(aes(colour=happy)) +
  geom_point()+
  facet_wrap(. ~ satfin, nrow = 3)+
  ylab("Percentage Happiness ")+
  scale_fill_manual(values=c("#19272e", "#c7a04e", "#39db7f"))+
  transition_reveal(year) 


animate(gif2, renderer = gifski_renderer(), fps = 60, height = 690, width =1000, duration = 3)
anim_save("gif2.gif")

```


This Plot illustrates that the percentage happiness changes with time and the financial and socio economic situation at a certain point of time in history till date. 
