---
title: "The Relationship between Financial and Job Satisfaction and Levels of Happiness Among Different Social Classes "
author: "Carrie Magee, Abhimanyu Barun, Ambar Pathak, and Kismat Khatri"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
#uploading packages
library(ezids)
library(dplyr)
library(ggplot2)
#library(DataExplorer)
library(plotly)
library(ggeasy)
library(ggpubr)
library(gganimate)
library(gifski)
library(transformr)
library(scales)
library(grid)
library(gridExtra)
library(grid)
library(gridExtra)
library(cowplot)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
gss1 <- data.frame(read.csv("../GSS.csv"))
#cleaning data
gss1_h = subset(gss1, happy == "Very happy" | happy == "Pretty happy" | happy == "Not too happy")
gss1_hsj = subset(gss1_h, satjob == "Very satisfied" | satjob == "Moderately satisfied" | satjob == "A little dissatisfied" | satjob == "Very dissatisfied")
gss1_hsjc = subset(gss1_hsj, class_ == "Lower class" | class_ == "Working class" | class_ == "Middle class" | class_ == "Upper class")
gss1_hsjcsf = subset(gss1_hsjc, satfin == "Pretty well satisfied" | satfin == "More or less satisfied" | satfin == "Not satisfied at all")
```


## Introduction

  There is still contention over whether money and success are the key to real happiness. Americans are often forced to sacrifice their happiness due to a culture that puts immense focus on establishing a successful career and working towards financial freedom. Furthermore, studies have shown the potential effects social class has on individual’s thoughts, feelings, and behaviors (Manstead, 2018). In relation, individuals of different social classes may have varying life experiences that can influence their financial and job satisfaction as well as their general happiness. 
  
  With the use of data from the General Social Survey, our report focuses on the general happiness of a individual in relation to their work satisfaction and financial satisfaction from 1980 to 2018. In order to measure job satisfaction, we used the statement *On the whole, how satisfied are you with the work you do..?* with responses ranging from *very satisfied, moderately satisfied, a little dissatisfied, or very dissatisfied*. Similarly, financial satisfaction was measured using the question *So far as you and your family are concerned, would you say that you are pretty well satisfied with your present financial situation, more or less satisfied, or not satisfied at all?* Our third variable "general happiness" was measured by asking *Taken all together, how would you say things are these days--would you say that you are very happy, pretty happy, or not too happy?*
  
  More specifically, we hoped to uncover any potential relationships between an individual's social class and their job satisfaction,financial satisfaction, and general happiness. Social class was measured by asking *If you were asked to use one of four names for your social class, which would you say you belong in?* with four responses including *the lower class, the working class, the middle class, or the upper class*. 
  

## General Social Survey dataset

### Descriptive Statistics
```{r}
par(mfrow=c(1,4))
pie_happy <- gss1_hsjcsf %>% count(happy) %>%
  arrange(desc(happy)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

pie_class <- gss1_hsjcsf %>% count(class_) %>%
  arrange(desc(class_)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

pie_satfin <- gss1_hsjcsf %>% count(satfin) %>%
  arrange(desc(satfin)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

pie_satjob <- gss1_hsjcsf %>% count(satjob) %>%
  arrange(desc(satjob)) %>% 
  mutate(percentages=n/sum(n) * 100, 
         ypos = cumsum(percentages)- 0.5*percentages )

plot1<-ggplot(pie_happy, aes(x="", y=percentages, fill=happy)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)+
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))+
  scale_fill_manual(values=c("#0290d1", "#e88738", "#39db7f", "#7f2bed"))+labs(title = "Happiness Variable",fill = "Happiness")+theme(plot.title = element_text(face="bold",hjust = 0.5))

plot2<-ggplot(pie_class, aes(x="", y=percentages, fill=class_)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)+
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))+
  scale_fill_manual(values=c("#0290d1", "#e88738", "#39db7f", "#7f2bed"))+labs(title = "Social Class Variable",fill = "Class")+theme(plot.title = element_text(face="bold",hjust = 0.5))

plot3<-ggplot(pie_satfin, aes(x="", y=percentages, fill=satfin)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)+
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))+
  scale_fill_manual(values=c("#0290d1", "#e88738", "#39db7f", "#7f2bed"))+labs(title = "Financial Satisfaction Variable",fill = "Financial Satisfaction")+theme(plot.title = element_text(face="bold",hjust = 0.2))

plot4<-ggplot(pie_satjob, aes(x="", y=percentages, fill=satjob)) +
geom_bar(stat="identity", width=1) + 
coord_polar("y", start=0)+
  geom_text(aes(x="", y = ypos, label = scales::percent(percentages, scale = 1)))+
  scale_fill_manual(values=c("#0290d1", "#e88738", "#39db7f", "#7f2bed"))+labs(title = "Job Satisfaction Variable",fill = "Job Satisfaction")+theme(plot.title = element_text(face="bold",hjust = 0.4))

plot_grid(plot1, plot2, plot3, plot4, align = "h", axis = "b", rel_widths = c(10, 10))
#grid.arrange(plot1, plot2, plot3, plot4, ncol=3)
```

The first pie chart shows that more than half of the people are pretty happy (57%) whereas the  individuals who report being very happy and not too happy make up 31% and 12% of the sample respectively.

The second  pie chart shows the makeup by different social classes for the sample. It shows that nearly half of the people fall under the working class (49.1%) and about 43% of the sample consider themselves to be middle class. Lower class individuals make up about 5% of the sample while upper class individuals make up about 3% of the sample. 
 
The third pie chart shows that nearly half of the people are more or less satisfied with their financial situation (~46%) whereas about 27% of people are pretty well satisfied and 27% are not satisfied at all with their financial situation.

The last pie chart shows about 48% of the sample are very satisfied with their job whereas 4.1% of people are dissatisfied with their job. The percentage of people who are a little dissatisfied and moderately satisfied are 10% and 38.2% respectively.

### Chi Squared Analysis
```{r}
sjPlot::tab_xtab(var.row = gss1_hsjcsf$happy, var.col =gss1_hsjcsf$satfin, title = "Level of Happiness by Financial Satisfaction Crosstabulation",var.labels = c("Happiness Level","Financial Satisfaction"), show.row.prc = TRUE,emph.total = TRUE)
```

The cross tabulation above shows the frequency of happiness levels by financial satisfaction. We wanted to further examine the relationship between financial satisfaction and happiness in order to see how feelings of financial satisfaction can potentially influence how happy an individual feels in life.  According to the chi-squared test of independence, there is a significant relationship between happiness and financial satisfaction (p<0.05). This means that an individual's level of happiness depends on their feelings of satisfaction with their current financial situation. More specifically, the table shows that a majority of individuals (54.5%) that report being not at all satisfied with their financial situation were not too happy in life. In addition, almost 49% of people that reported being more or less satisfied with their financial situation were pretty happy with their life while about 42% of individuals that were pretty well satisfied with their financial situation reported being very happy in life. The results clearly imply that the more financially satisfied an individual is, the happier they are in life. 

```{r fig.height = 8, fig.width = 13}
sjPlot::tab_xtab(var.row = gss1_hsjcsf$happy, var.col = gss1_hsjcsf$satjob,title = "Level of Happiness by Job Satisfaction Crosstabulation",var.labels = c("Happiness Level","Job Satisfaction"), show.row.prc = TRUE,emph.total = TRUE)
```

The cross tabulation above shows the frequency of happiness levels by job satisfaction. We wanted to further examine the relationship between job satisfaction and happiness in order to see how job satisfaction can potentially influence how happy an individual feels in life. According to the chi-squared test of independence, there is a significant relationship between job satisfaction and happiness (p<0.05). This means that an individual's level of happiness depends on their satisfaction with their current job. The table shows that individuals that a majority of individuals that reported being very dissatisfied (12.2%) or a little dissatisfied (19.8%) with their jobs were not too happy in their lives. About 44% of people that were moderately satisfied with their jobs reported being pretty happy in life while 65.9% of people who were very satisfied with their jobs reported being very happy in life. Considering that individuals who reported being more dissatisfied with job also reported being unhappy and those who reported being more satisfied were more happy shows  that job satisfaction may play a crucial role in how happy individuals are in life. 

### Graph 1
```{r fig.height = 7, fig.width = 12}
# making plot of relationship between class and financial satisfaction
ggplot(gss1_hsjcsf, aes(x=satfin, fill=happy)) +
  geom_bar(position = "dodge")+
  facet_wrap(~class_)  +
  labs(title="Relationship Between Financial Satisfaction and Happiness Levels by Social Class",x="Financial Satisfaction", y = "Frequency",fill="Happiness")+
  theme_bw(base_size=14)+theme(plot.title = element_text(face="bold",hjust = 0.5),axis.text.x = element_text(angle=30, vjust = 1, hjust = 1))+
  scale_fill_manual(values=c("#0290d1", "#e88738", "#39db7f", "#7f2bed"))+scale_x_discrete(limits = c("  Not satisfied at all","More or less satisfied","Pretty well satisfied"))

```

This figure illustrates the relationship between individuals' financial satisfaction and general happiness by social class. There is also a disproportionate number of individuals in the middle and working class groups in comparison to the lower and upper class individuals. For lower, working, and middle class groups, a majority of individuals were more or less satisfied with their financial situation and reported feeling pretty happy in their life. It is important to note that there is a similar trend in the middle class group for both job satisfaction and financial satisfaction in relation to happiness. Middle class individuals that are highly satisfied with their job or financial situation were equally high in both the pretty happy and very happy responses.For upper class individuals, a large number were pretty well satisfied with their financial situation and reported being very happy in life. As we can see, the results show that the higher the social class the more you become satisfied with your financial situation and the more you report being very happy while the more you go down in class, the more your satisfaction and happiness become average (as in more or less satisfied and pretty happy).

### Graph 2
```{r fig.height = 7, fig.width = 12}
# making plot of relationship between class and job satisfaction
ggplot(gss1_hsjcsf, aes(x=satjob, fill=happy)) +
  geom_bar(position = "dodge")+
  facet_wrap(~class_)  +
  labs(title="Relationship Between Job Satisfaction and Happiness Levels by Social Class",x="Job Satisfaction", y = "Frequency",fill="Happiness")+
  theme_bw(base_size = 14)+theme(plot.title = element_text(face="bold",hjust = 0.5),axis.text.x = element_text(angle=30, vjust = 1, hjust = 1))+
  scale_fill_manual(values=c("#0290d1", "#e88738", "#39db7f", "#7f2bed"))+scale_x_discrete(limits = c("Very dissatisfied","A little dissatisfied", "Moderately satisfied","Very satisfied"))

```

The figure above looks at the relationship between individuals' job satisfaction and general happiness by social class. For context, there is a larger number of working and middle class individuals than upper and lower class individuals. The results show that many lower class individuals were very satisfied with their job and that those same individuals were pretty happy overall. A majority of working class individuals were also very satisfied with their job and were also pretty happy. Interestingly, many middle class individuals were very satisfied with their jobs but reported being equally pretty happy and very happy in their lives. The upper class individuals were the only group that had more very happy individuals with very high job satisfaction. In other words, most individuals within the four social class groups were very satisfied with their jobs but only upper class individuals had more very happy individuals compared to pretty happy individuals. These results provide some insight into how social class may have a greater influence on individual's feelings of financial satisfaction and thus also influencing their feelings of happiness in life.  


### Time-Series Analysis

We created one subset of dataframe by adding a column of  percentage valueof every possible combination of happiness, year and job satisfaction and similarly another subset for that showed percentage value of every possible combination of happiness, year and financial satisfaction.

```{r fig.height = 7, fig.width = 10}

# Calculations prior to plotting happiness levels compared to the job satisfaction over the course of 50 Years.

gss_satjob <- gss1_hsjcsf %>% 
  group_by(year, happy, satjob) %>% 
  summarise(total_count=n(),
            .groups = 'drop') %>% 
  as.data.frame()

gss_satjob_perc <- transform(gss_satjob,
                             perc = ave(total_count,
                                        year,
                                        FUN = prop.table))

# Plot to analyze happiness levels compared to the job satisfaction over the course of 50 Years.

ggplotly(gss_satjob_perc %>%
           ggplot( aes(x=year, y=perc)) +
           geom_line(aes(colour=happy)) +
           geom_point(size=0.5)+
           xlab("Years")+
           ylab("Percentage Happiness")+
           labs(title="Job Satisfaction Levels")+
           theme(plot.title=element_text(hjust=0.5, size = 10))+
           facet_wrap(. ~ satjob, nrow = 4)+
           guides(color = guide_legend(title = "Happiness Type"))+
           transition_reveal(year)+
           scale_color_manual(values=c("#0290d1", "#e88738", "#39db7f")))


gif1 <- gss_satjob_perc %>%
           ggplot( aes(x=year, y=perc)) +
           geom_line(aes(colour=happy)) +
           geom_point(size=0.5)+
           xlab("Years")+
           ylab("Percentage Happiness")+
           labs(title="Job Satisfaction Levels")+
           theme(plot.title=element_text(hjust=0.5, size = 10))+
           facet_wrap(. ~ satjob, nrow = 4)+
           guides(color = guide_legend(title = "Happiness Type"))+
           transition_reveal(year)+
           scale_color_manual(values=c("#0290d1", "#e88738", "#39db7f"))


#animate(gif1, renderer = gifski_renderer(), fps = 60, height = 690, width =1000, duration = 3)
#anim_save("gif1.gif")

```

The graph tells us a story of how job satisfaction relates to the level of happiness, over time, of the surveyed individuals, with time being on the x-axis, the percentage of happiness on the y-axis (with its subcategory description in the legends that are right aligned), and the job satisfaction in a matrix of panels where each panel represents one subcategory of job satisfaction. It is visible that a little amount of discomfort is imminent in human lives, which is exactly what the data tells us when we see that there is a group of people who are a little dissatisfied irrespective of the amount of happiness that they tend to experience in life. When it comes to being moderately satisfied by their job, we can see that this sub-metric is directly proportional to being moderately happy in life. It is also evident that the people who are very dissatisfied with their jobs tend to have very low levels of happiness as opposed to the people who are satisfied with their jobs, where the percentage happiness number is relatively high for pretty (moderately) happy and/or very happy people. If we add on to this by studying it based off of the x-axis, which is time, we can say that we see a sudden upward movement from 1998 to 2000 for the percentage value of pretty happy and very happy people, which is exactly where the professional and business services industry took off (as per the United States Bureau of Labor Statistics) because of feature-rich computers that had been commercialized and reliable technology came into being, which naturally meant a decline in the not-so-happy percentage. This is important in proving the validity of the data because moderately satisfied people are usually the ones who are earning just enough to not suffer but not enough to call themselves rich, and the sectoral shift from the manufacturing-heavy job sector to the professional and business services sector meant more job opportunities for the middle class families. The second major movement in the graph worth pointing out is when the global pandemic hit the United States. We see a symmetry in the not-very-happy percentage going up while the happiness percentage values are coming down irrespective, especially amongst the people who were very satisfied with their job as the country came to a stop due to a lockdown and people lost their job in the process.



```{r fig.height = 7, fig.width = 10}

# Calculations prior to plotting happiness levels compared to the financial satisfaction over the course of 50 Years.

gss_satfin <- gss1_hsjcsf %>% 
  group_by(year, happy, satfin) %>% 
  summarise(total_count=n(),
            .groups = 'drop') %>% 
  as.data.frame()

gss_satfin_perc <- transform(gss_satfin,
                             perc = ave(total_count,
                                        year,
                                        FUN = prop.table))

# Plot to analyze happiness levels compared to the financial satisfaction over the course of 50 Years.

ggplotly(gss_satfin_perc %>%
           ggplot( aes(x=year, y=perc)) +
           geom_line(aes(colour=happy)) +
           geom_point(size=0.5)+
           facet_wrap(. ~ satfin, nrow = 3)+
           xlab("Years")+
           ylab("Percentage Happiness ")+
           labs(title="Financial Satisfaction Levels ")+
           theme(plot.title=element_text(hjust=0.5, size = 10))+
           facet_wrap(. ~ satfin, nrow = 3)+
           guides(color = guide_legend(title = "Happiness Type"))+
           transition_reveal(year)+
           scale_color_manual(values=c("#0290d1", "#e88738", "#39db7f")))

gif2<-gss_satfin_perc %>%
  ggplot( aes(x=year, y=perc)) +
  geom_line(aes(colour=happy)) +
  geom_point(size=0.5)+
  facet_wrap(. ~ satfin, nrow = 3)+
  xlab("Years")+
  ylab("Percentage Happiness ")+
  labs(title="Financial Satisfaction Levels ")+
  theme(plot.title=element_text(hjust=0.5, size = 10))+
  facet_wrap(. ~ satfin, nrow = 3)+
  guides(color = guide_legend(title = "Happiness Type"))+
  transition_reveal(year)+
  scale_color_manual(values=c("#0290d1", "#e88738", "#39db7f"))


animate(gif2, renderer = gifski_renderer(), fps = 60, height = 690, width =1000, duration = 3)
anim_save("gif2.gif")

```

This plot illustrates the relationship between financial satisfaction and level of happiness over time for the surveyed individuals, with time on the x-axis, percentage happiness on the y-axis (with its subcategory description in the legends that are right aligned), and financial satisfaction in a matrix of panels where each panel represents one subcategory of job satisfaction. It is visible that when we stack a big picture metric like financial satisfaction against job satisfaction, and inside, if we look at people who are more or less satisfied with their financial status, we find that there are a huge percentage of people who are moderately happy and even very happy, with the not too happy percentage being on the low until COVID-19 hits the country, after which the not too happy value goes up, bringing down the very happy percentage equally dramatically. The people who are moderately happy in this category have a high number, which is directly proportional to their being partially satisfied, and this value is contrary to the not too happy and very happy values as it goes up, which can be backed up by the fact that these people are generally in the middle class or small business category, and COVID-19 gave them a huge opportunity to be profitable from their small or local businesses while all the big firms were shut down.When we move on to people not satisfied at all, we see that the happiness percentage never broke the 20 percent barrier and stays below that even after a gradual rise from 11.95 percent in 1972 and hitting the barrier three times in 1983, 1993, and 2010, respectively. The very happy number, as expected, stays at the lowest of the lows for not financially satisfied people, and the not happy percentage very slowly goes up with the rise in inflation over all these years with a steep upward movement, and this value represents the people already having financial problems, which rose after COVID-19. This is justified by the very happy number going further down along with the moderately happy number going down. If we move on to the last part of our graph, we find both happiness values pretty higher than the not too happy value, which makes sense considering the group we are highlighting right now is pretty well satisfied, and as the inflation rate went up along with time, we saw a slow and steady rise in the unhappiness percentage with a more steep climb from 2018 to 2020, which effectively caused a sharp fall for the very happy value, even for the pretty well satisfied group of people, which can be justified by the combined negative effect of inflation and COVID-19 affecting jobs and businesses alike.